---
title: "Data_statistical analysis"
author: "Jingxin Yuan"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    latex_engine: xelatex
    fig_crop: false
    toc: true
    toc_depth: 1
    number_sections: true
bibliography: references.bib
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Introduction
## Research question:

Does gender influence Chronic kidney disease (CKD) mortality rate among adults aged 60+ years in China?

## Scientific plausibility and context:

Chronic kidney disease (CKD) is an important public health problem in China. A recent study of CKD burden in China shows that CKD contributes substantially to national disease burden, especially among olders [@li2023ckj]. A study points out that gender affect CKD prevalence, disease patterns, treatment decisions, and clinical outcomes [@mayne2023sex]. Evidence from China CDC Weekly shows that although fewer males were affected by hypertension-related CKD, the number of deaths was higher among males than females [@bu2024htnckdchina]. Above all, these findings support the plausibility of studying whether CKD mortality differs by gender in China.

# Material & Methods
## Data

The data used in this project is derived from the Global Burden of Disease(GBD) 2023 databases for 60+ years old people in China from 2013 to 2023. This dataset includes CKD prevalence, DALYs and death rate, stratified by year, sex and age group. The complete data set and accompanying information are available at [vizhub.healthdata.org/gbd-results](https://vizhub.healthdata.org/gbd-results/)

Data cleaning steps included:
- Removal of missing values
- Detection and replacement of outliers using the IQR method
- Subsetting relevant variables (years, sex, mortality rate)

```{r}

# Packages
library(tidyverse)
library(knitr)

# Set working directory
setwd("C:/Users/乌乌没有仙人/Desktop/6270/HW5/VTPEH6270_Jingxin-Yuan/Data/Original data")

# Load file
data_ckd = read.csv("IHME-GBD_CKD_20260130.csv")

#Check if it contains NA
n_na <- sum(is.na(data_ckd))
cat("The amount of NA is",n_na,end='\n')

#Check available years
sort(unique(data_ckd$year))

#Check of Abnormal Value
#Define IQR measure to detect abnormal value
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  return(x < lower_bound | x > upper_bound)
}

#Detected column
numeric_cols <- c("val", "upper", "lower")

#Detect
outliers_df <- data_ckd %>%
  mutate(across(all_of(numeric_cols), detect_outliers))

outlier_summary <- colSums(outliers_df[numeric_cols], na.rm = TRUE)
print(outlier_summary)

#Replace abnormal value with median value
median_vals <- apply(data_ckd[numeric_cols], 2, median, na.rm = TRUE)

data_ckd[numeric_cols] <- lapply(numeric_cols, function(col) {
  data_ckd[[col]] <- ifelse(outliers_df[[col]], median_vals[col], data_ckd[[col]])
})

# filter data
data_ckd_old <- data_ckd %>%
  filter(
    location_name == "China",
    age_name == "60+ years",
    measure_name == "Deaths"
  ) %>%
  select(year, sex_name, val)

# Examine data structure
str(data_ckd_old)

# Create variable description table
var_table <- tibble(
  Variable_Name = c("year", "sex_name", "val"),
  Variable_Type = c("Discrete", "Categorical", "Continuous"),
  R_Class = c(class(data_ckd_old$year),
              class(data_ckd_old$sex_name),
              class(data_ckd_old$val)),
  Description = c(
    "Observation year",
    "Gender (Male or Female)",
    "CKD mortality rate per 100,000 population"
  )
)
var_table

```

## Statistical Analyses
### Hypothesis

- Null hypothesis (H0): There is no difference in CKD mortality rates between males and females.
- Alternative hypothesis (H1): CKD mortality rates differ between males and females.

### Approach

In this study, gender is treated as the explanatory variable, which is a categorical variable with two groups (male and female), while CKD mortality rate is the outcome variable and is measured on a continuous scale. Because the objective is to compare the average mortality rates between two independent groups, an independent two-sample t-test is selected as the appropriate statistical method.

### Assumption Check

Normality of the outcome variable within each group was assessed using the Shapiro-Wilk test and QQ plots.

```{r}

# Split data
male_data <- data_ckd_old %>% filter(sex_name == "Male") %>% pull(val)
female_data <- data_ckd_old %>% filter(sex_name == "Female") %>% pull(val)

# Shapiro test
shapiro.test(male_data)
shapiro.test(female_data)

# QQ plots
qqnorm(male_data, main = "QQ Plot for Male")
qqline(male_data)
qqnorm(female_data, main = "QQ Plot for Female")
qqline(female_data)

```

The Shapiro-Wilk test suggested that mortality rates in both males (p = 0.2769) and females (p = 0.2076) don't significantly deviate from normality. The QQ plots also show that most points lie close to the reference line, with slight deviations at the tails. Overall, the normality assumption appears reasonable, supporting the use of a t-test. In addition to normality, the t-test also assumes that observations are independent and that the two groups have similar variability. In this dataset, the values are reported by year, so there could be some temporal pattern, but for this analysis they are treated as approximately independent. The variability between males and females is also fairly similar, and a Welch two-sample t-test is used, which does not require equal variances. Taken together, these assumptions seem reasonable for this comparison.

### Code

All code used for data cleaning, analysis and visualization is available in the GitHub repository linked below, allowing full reproducibility of the results.

GitHub repository: https://github.com/JingxinYuan/VTPEH6270_Jingxin-Yuan.git

```{r}

t_test_result <- t.test(val ~ sex_name, data = data_ckd_old)
t_test_result

```

# Results
## Visualization

```{r}

p3 <- ggplot(data_ckd_old, aes(x = year, y = val, color = sex_name)) +
  geom_line(size = 1.2) +
  labs(
    title = "CKD Mortality Rate in China (Age 60+ Years)",
    x = "Year",
    y = "Mortality Rate per 100,000",
    color = "Gender"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )
ggsave("C:/Users/乌乌没有仙人/Desktop/6270/HW5/VTPEH6270_Jingxin-Yuan/Output/Figures/Gender trend on mortality of CKD of 60+ years people in China.png",plot = p3)

# Show plot
p3

```

This time trend visualization shows that CKD mortality rates among adults aged 60+ years in China declined from the 2013 to 2023, followed by a slight increase in subsequent years. Across the entire study period, mortality rates remain consistently higher in males than in females. The parallel movement of the two lines suggests that both genders experienced similar temporal patterns, but males maintained a persistent higher mortality burden. Overall, the figure supports the hypothesis of a sustained gender disparity in CKD mortality.

## Statistical Results

The independent two-sample t-test shows that the mean CKD mortality rate is 43.30 for females and 46.11 for males, indicating a higher average mortality among males. However, this difference isn't statistically significant since t-value is -1.94 and p-value is 0.0667. The 95% confidence interval (-5.83 to 0.21) includes zero, suggesting that the observed difference may be due to random variation rather than a true effect. Although the result isn't statistically significant at the 0.05 level, the p-value suggests a marginal trend.

# Conclusion

The independent two-sample t-test shows that males have higher mean CKD mortality rates than females. However, this difference isn't statistically significant at the 0.05 level since p-value is 0.0667, and therefore the null hypothesis can't be rejected. 
Although the result doesn't provide sufficient evidence of a gender difference, the p-value is close to the significance threshold and may suggest a potential trend. This raises the possibility that the non-significant finding could be influenced by factors such as limited sample size or variability in the data, which will be further explored in the simulation analysis.

# AI Use Disclosure Statement

*As part of this assignment, please indicate whether you used any AI-based tools (e.g., ChatGPT, Claude, Copilot, Gemini, etc.). Indicate:*

* *Did you use AI? Yes / No*
  Yes

* *If yes: Write a short disclosure statement (2–3 sentences) describing:*
  Artificial intelligence tools were used to assist with code development for this assignment. All analyses, results, and interpretations were reviewed and verified by the author.

* *Which tool(s) you used.*
  ChatGPT

* *How you used it (e.g., to help decide on an analysis approach, to generate a first draft of code, to improve or debug code you had written yourself, etc.).*

  *Example: “This document was generated using Claude to generate original code, which was then reviewed and adjusted” or "This document was generate using ChatGPT to assist with debugging of code generated by the author".*
  
  This assignment was generated using ChatGPT to find corresponding coding packages and debug the code generated by the author.
  