---
title: "Data_exploration"
author: "Jingxin Yuan"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    latex_engine: xelatex
    fig_crop: false
    toc: true
    toc_depth: 1
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Introduction

#Background
Chronic Kidney Disease(CKD) is a growing global challenge, driven by aging populations, environmental factors and lifestyle changes. In China, efforts in early detection and prevention face significant borders, especially for olders. Furthermore, current data on CKD's national burden in China is outdated. There is a need for up-to-date and evidence-based analysis to understand recent and specific trends of CKD in China. 

#Data set overview
This project uses data from the Global Burden of Disease(GBD) 2023 databases for 60+ years old people in China from 2013 to 2023. This dataset includes CKD prevalence, DALYs and death rate, stratified by year, sex and age group. The complete data set and accompanying information are available at [vizhub.healthdata.org/gbd-results](https://vizhub.healthdata.org/gbd-results/)


```{r,results='hide'}

# Packages
library(tidyverse)
library(scales)
library(dplyr)
library(ggplot2)
library(knitr)

```

# Load Data
```{r}

# Set working directory
setwd("C:/Users/乌乌没有仙人/Desktop/6270/HW5/VTPEH6270_Jingxin-Yuan/Data/Original data")

# Load file
data_ckd = read.csv("IHME-GBD_CKD_20260130.csv")

# Preview the data
head(data_ckd)

```

# Data cleaning

## Check of NA
```{r}

# Check of NA
n_na <- sum(is.na(data_ckd))
cat("The amount of NA is",n_na,end='\n')

# Check available years
sort(unique(data_ckd$year))

```


## Check of Abnormal Value
```{r}

# Define IQR measure to detect abnormal value
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  return(x < lower_bound | x > upper_bound)
}

# Detected column
numeric_cols <- c("val", "upper", "lower")

# Detect
outliers_df <- data_ckd %>%
  mutate(across(all_of(numeric_cols), detect_outliers))

outlier_summary <- colSums(outliers_df[numeric_cols], na.rm = TRUE)
print(outlier_summary)

# Replace abnormal value with median value
median_vals <- apply(data_ckd[numeric_cols], 2, median, na.rm = TRUE)

data_ckd[numeric_cols] <- lapply(numeric_cols, function(col) {
	  data_ckd[[col]] <- ifelse(outliers_df[[col]], median_vals[col], data_ckd[[col]])
	})

# Output processed data
write.csv(data_ckd, "C:/Users/乌乌没有仙人/Desktop/6270/HW5/VTPEH6270_Jingxin-Yuan/Data/Processed data/IHME-GBD_CKD_Processed.csv")

```

# Variable Overview

## Mutate class of important data column
```{r}

data_ckd = data_ckd %>%
  mutate(
    year = as.integer(year),
    val = as.numeric(val),
    lower = as.numeric(lower),
    upper = as.numeric(upper)
  )

```

## Create a variable table
```{r}

data_variable = names(data_ckd)
data_class = sapply(data_ckd,class)
data_type = sapply(data_ckd, function(x){
  if (is.numeric(x)) {
    if (all(x %% 1 == 0, na.rm = TRUE)) "discrete"
    else "continuous"
  } else {
    "categorical"
  }
})

# Combine to one table
data_ckd_table = data.frame(
  variable = data_variable,
  class = data_class,
  type = data_type
)

# Output 10 head rows of variable table
data_ckd_table = head(data_ckd_table,10)
kable(data_ckd_table, caption = "Table 1. Variable overvie (first 10 variables)")

```

# Distribution Checking

*Histogram on the burden of CKD of 60+ years people in China*
```{r, results='hide'}

# Create Scubset
unique(data_ckd$measure_name)

data_ckd_plot = data_ckd %>%
  filter(
    location_name == "China",
    age_name == "60+ years",
    measure_name == "DALYs (Disability-Adjusted Life Years)"
  )

# Histogram
mean_d <- mean(data_ckd_plot$val, na.rm = TRUE)
median_d <- median(data_ckd_plot$val, na.rm = TRUE)

vlines <- data.frame(
  stat = c("Mean", "Median"),
  xint = c(mean_d, median_d)
)

p1 <- ggplot(data_ckd_plot, aes(x = val)) +
  geom_histogram(bins = 25, fill = "grey") +
  geom_vline(
    data = vlines,
    aes(xintercept = xint, linetype = stat)
  ) +
  scale_linetype_manual(values = c("Mean" = "dashed", "Median" = "dotted")) +
  theme_classic() +
  labs(
    title = "Distribution of CKD DALYs Rate of\n 60+ Years Old People in China (2013–2023)",
    x = "DALYs rate (per 100,000)",
    y = "Count",
    linetype = "Statistic",
    caption = "Figure 1. Histogram of CKD DALYs rate in China with mean and median lines."
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
ggsave("C:/Users/乌乌没有仙人/Desktop/6270/HW5/VTPEH6270_Jingxin-Yuan/Output/Figures/Distribution on the burden of CKD of 60+ years people in China.png",plot = p1)

# Show plot
p1

```
*DALYs rate is the representative variable of disease burden, and the figure shows it generally range from 1000 to 1300 per 100,000 population with most values clustered in the middle. The distribution is lightly right-skewed, indicating several years with relatively higher DALYs rate among 60+ years old people in China.*

# Broad trend exploration

*Scatterplot on the burden of CKD of 60+ years people in China*
```{r}

#Create subset
data_ckd_scatterplot <- data_ckd %>%
  filter(
    location_name == "China",
    age_name == "60+ years",
    measure_name %in% c("Prevalence","DALYs (Disability-Adjusted Life Years)","Deaths")
    )

# Scatterplot on the burden of CKD of 60+ years people in China
p2 <- ggplot(data_ckd_scatterplot, aes(x = year, y = val, color = sex_name)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~ measure_name, ncol = 1, scales = "free_y") +
  theme_minimal() +
  scale_x_continuous(breaks = 2013:2023) +
  labs(
    title = "Trends in CKD Prevalence, DALYs and Death Rate of \n 60+ Years Old People in China (2013–2023)",
    x = "Year",
    y = "Rate (per 100,000)",
    color = "Sex",
    caption = "Figure 2. Trends of CKD prevalence, DALYs and death rates of \n 60+ Years Old People in China from 2013 to 2023, stratified \n by sex. Shaded areas represent the 95% confidence interval."
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
ggsave("C:/Users/乌乌没有仙人/Desktop/6270/HW5/VTPEH6270_Jingxin-Yuan/Output/Figures/Scatterplot on the burden of CKD of 60+ years people in China.png",plot = p2)

# Show plot
p2

```
*The figure shows a U-shaped pattern on DALYs and death rates of CKD among people aged 60+ in China from 2013 to 2023, which declined from 2016 to 2020 and then rising again for both males and females. Males consistently exhibit higher DALYs and death rates than females across the entire period, indicating a greater disease burden among older men. But CKD prevalence remains nearly constant over years for both sexes, suggesting that while the number of people living with CKD stayed stable, the severity of health loss and mortality fluctuated.*

# Focused subset for mortality and gender

## Form subset
```{r}

# Filter 60+ years old adults
data_ckd_old <- data_ckd %>%
  filter(
    location_name == "China",
    age_name == "60+ years",
    measure_name == "Deaths"
  ) %>%
  select(year,sex_name,val)

```


## Grouped Summary Statistics
```{r}

summary_table <- data_ckd_old %>%
  group_by(sex_name) %>%
  summarise(
    Mean = mean(val),
    SD = sd(val),
    Median = median(val),
    Min = min(val),
    Max = max(val)
  )
kable(summary_table, digits = 2)

```
*Descriptive statistics show that CKD mortality rates among adults aged 60+ years in China are consistently higher in males than in females. The mean mortality rate is 46.11 per 100,000 in males compared to 43.30 in females. Similarly, the median value is higher in males (46.42) than females (43.48), suggesting that the difference is not driven by extreme values. The relatively similar standard deviations which are 3.28 and 3.50 show comparable variability between the two groups. Overall, these statistics suggest a higher mortality burden among males.*

## Mortality difference
```{r}

data_ckd_old_diff <- data_ckd_old %>%
  pivot_wider(names_from = sex_name, values_from = val) %>%
  mutate(
    mortality_difference = Male - Female
  ) %>%
  select(year, mortality_difference)

data_ckd_old <- data_ckd_old %>%
  left_join(data_ckd_old_diff, by = "year")
head(data_ckd_old)

```
*This study creates a derived variable "mortality_difference" to show the difference in mortality rates between males and females. Positive values mean higher mortality among males, while negative values mean higher mortality among females.*

## Comparative visualization on mortality
```{r}

p3 <- ggplot(data_ckd_old, aes(x = year, y = val, color = sex_name)) +
  geom_line(size = 1.2) +
  labs(
    title = "CKD Mortality Rate in China (Age 60+ Years)",
    x = "Year",
    y = "Mortality Rate per 100,000",
    color = "Gender"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )
ggsave("C:/Users/乌乌没有仙人/Desktop/6270/HW5/VTPEH6270_Jingxin-Yuan/Output/Figures/Gender trend on mortality of CKD of 60+ years people in China.png",plot = p3)

# Show plot
p3

```
*This time trend visualization shows that CKD mortality rates among adults aged 60+ years in China declined from the 2013 to 2023, followed by a slight increase in subsequent years. Across the entire study period, mortality rates remain consistently higher in males than in females. The parallel movement of the two lines suggests that both genders experienced similar temporal patterns, but males maintained a persistent higher mortality burden. Overall, the figure supports the hypothesis of a sustained gender disparity in CKD mortality.*

## Visualize mortality difference between male and female
```{r}

p4 <- ggplot(data_ckd_old, aes(x = year, y = mortality_difference)) +
  geom_col(fill = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Sex Difference in CKD Mortality (Male − Female)",
    x = "Year",
    y = "Mortality Rate Difference (per 100,000)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
ggsave("C:/Users/乌乌没有仙人/Desktop/6270/HW5/VTPEH6270_Jingxin-Yuan/Output/Figures/Mortality difference between gender of CKD of 60+ years people in China.png",plot = p4)

# Show plot
p4

```
*The bar chart shows the annual difference in CKD mortality rates between males and females aged 60+ years in China. Across all observed years, the difference remains above zero, indicating consistently higher mortality among males. The magnitude of the difference fluctuates over time, with a gradual increase from 2010 and a peak in 2022, suggesting a temporary widening of the gender disparity. Although the gap narrows slightly in the most recent year, the persistent positive difference supports the presence of higher male mortality.*
